---
title: "Longitudinal individualisation with LibeRator"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Longitudinal individualisation with LibeRator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LibeRator is one product designed toward eventual clinical use. Its current
**Research** label communicates that clinical validation is not yet complete;
it does not identify a separate research-only edition.

LibeRator separates four things that are often mixed together in a dosing script:

1. immutable patient evidence;
2. a qualified population PK/PD model;
3. a versioned therapeutic endpoint;
4. an assessment or regimen comparison tied to exact hashes of the first three.

This separation makes it possible to revisit what was known at each point in a patient's timeline and to distinguish a changing posterior estimate from an explicitly time-varying latent parameter.

## Patient evidence and missing covariates

```{r example}
library(LibeRator)
example <- lator_example_aed()
example$patient

lator_covariate_at(
  example$patient, "WT", times = c(0, 168, 180),
  method = "locf", max_age = 365 * 24
)
```

The weight scheduled at hour 168 is explicitly missing. LOCF can still be an appropriate modelling policy, but the resolved value carries the `resolved_after_missing` status, source event, age, and missing reason. A population fallback is used only when the caller supplies it explicitly.

Medication records are added to the treatment profile before entering doses or
TDM. Optional monitored analytes keep active metabolites linked to their parent
treatment:

```{r medication, eval=FALSE}
patient <- lator_patient_medication_add(
  example$patient, "carbamazepine", "antiseizure",
  monitoring_analytes = "carbamazepine-10,11-epoxide"
)
```

Model covariates named `COMED_<DRUG>`, `TRT_<DRUG>`, or `DDI_<DRUG>` are
resolved from these active treatment profiles. This activates an interaction
effect only when the model explicitly defines one; LibeRator does not infer an
effect size or mechanism from the medication list. Explicit effective-dated
covariate records with the same name override the profile fallback, supporting
interaction start/stop states along the patient timeline.

## Static and dynamic patient states

```{r assessment}
static <- lator_assess(
  example$patient, example$model, example$endpoint,
  covariate_policies = list(WT = list(method = "locf", max_age = 365 * 24))
)

dynamic <- lator_assess(
  example$patient, example$model, example$endpoint,
  mode = "dynamic", process_scale = 0.1,
  covariate_policies = list(WT = list(method = "locf", max_age = 365 * 24))
)

dynamic$eta_trajectory
```

Static mode estimates one ETA vector from all visible observations. Dynamic mode turns each ETA into an occasion-specific state. The joint prior is a Gaussian random walk: adjacent states are correlated, their difference has an explicit process covariance, and PK compartment amounts remain continuous at boundaries. Add `state_boundary` events only where the scientific design supports a possible change; adding a state at every sample is usually weakly identified.

## Endpoint-based regimen comparison

```{r regimen, eval=FALSE}
candidates <- lator_regimen_candidates(
  amounts = c(100, 200, 300), intervals = c(12, 24), horizon = 168
)
comparison <- lator_regimen_optimise(
  dynamic, example$patient, candidates,
  endpoint = example$endpoint, nsim = 500, seed = 42
)
comparison$summary

# Promote deliberately selected candidates to separate forecast artifacts.
selected <- comparison$summary$candidate_id[1:2]
forecasts <- lapply(
  selected,
  function(candidate_id) lator_regimen_predict(comparison, candidate_id)
)
forecasts[[1L]]$forecast
```

Each posterior draw replays the complete dosing history before the future candidate regimen. For the AED endpoint, the default score is distance from the midpoint of the supplied therapeutic range. Other endpoint objects can evaluate pre-transplant windows, AUC/trough ranges, or fraction of time above a longitudinal MIC.

The optimisation step already generates candidate trajectories efficiently.
`lator_regimen_predict()` does not silently re-fit or re-simulate them; it
creates linked forecast artifacts from the selected candidates and retains
their uncertainty, regimens, assessment identifier, and endpoint evaluations.
The GUI stacks selected forecast plots vertically for comparison. This keeps
the transition from ranked options to future prediction explicit and reviewable.

## Graphical workbench

```{r gui, eval=FALSE}
workspace <- lator_workspace(
  "~/LibeR/liberator-workspace",
  passphrase = "use a long, unique research passphrase"
)
lator_patient_save(workspace, example$patient)
lator_gui(
  workspace = workspace,
  models = list(aed = example$model),
  endpoints = list(aed = example$endpoint),
  library_root = "~/LibeR-data/library"
)
```

The **Find best model** action evaluates only current, scoped clinical-use
qualifications from LibeRary and fails closed when no model is suitable or when
the leading candidates are too close to distinguish safely. The workbench
retains the complete patient-context snapshot, qualification identifier, model
hash, score components, blockers, and selection criteria. Clinical deployment
of this same product still needs the additional model governance, identity,
authorization, safety, independent validation, and operational controls
described in `SECURITY.md`.
